Localization properties, observable character and computational reliability of the linear response in open systems

### 2018-06-13 AnalyticStructure_main.tex
continuing

It is easy to see from these concepts that a computational treatment that is tailored for the discretization of bound states
will not be effective for continuum states, due to the different long-range asymptotic behaviour - which in turn is a clear consequence of the 
BC of the problem. In a numerical treatment, states with $\epsilon > 0$ will have a value of their energy that will \emph{always}
depend on the computational setup.

A direct numerical evidence of this problem has been discussed in \cite{boffi2016}. The authors of this paper analyze the effects of a finite basis set on the 
occupied and virtual orbitals of a molecular system and establish a clear correspondence between the orbital localization character, and the independence of its properties (\emph{e.g.} energy) 
with respect to the basis set. It is shown that (localized) orbitals with a bound-state behavior in the long range constitute a genuine discrete set and a finite basis is 
able to provide a reliable description of these states if the sampled domain contains their natural sizes. On the contrary, unbound (delocalized) orbitals form a continuum of states in the 
true system with open boundaries. For these class of objects the presence of a finite computational domain has a deeper impact since introduces a fictitious discretization of the energy levels. 
Furthermore, the energies of the unbound orbitals are not stable with respect to a change of the size of the computational domain and approach to zero as long as it is increased.

@@

In the contest of the LR the response density satisfies an equation of motion written in the form of a quantum Liouville operator (for example see \cite{baroni2008})
\be\lb{LiouvillianRhopomegaDef1}
\left(\omega - \Liouv\right) \dm'_\Phi(\omega) =  \commutator{\op\Phi(\omega)}{\dmnot} \;,
\ee
where $\op\Phi(\omega)$ represents the perturbing field. 
The inspection of the linear order time evolution of the eigenstates of $\hnot$ evidence that the response density acts as a transition 
operator, linking the occupied and empty subspaces of $\hnot$, and satisfies the \emph{transverse} condition  
\be\lb{RhopTransverseDef1}
\dm'_\Phi = \dm_{\perp} \doteq \dmnot\dm'_\Phi\op Q_0 + \op Q_0\dm'_\Phi\dmnot \;,
\ee
where $\op Q_0=\identity-\dmnot$ is the projector in the subspace orthogonal to the occupied states of $\hnot$. 
Due to this fact we can introduce an explicit representation of the 
response density parametrized as
\be\lb{rhoPrimeFluctuationStateDef1}
\dm'_\Phi(\omega) = \sum_p\left(\ketbra{\psi_p}{f_p^\Phi(-\omega)} + \ketbra{f_p^\Phi(\omega)}{\psi_p}\right) \;.
\ee
Here we have introduced the set of $\omega$-dependent \emph{fluctuation states} $\{\ket{f_p^\Phi(\omega)}\}$, 
that have values in the \emph{unoccupied} subspace of $\hnot$. 
An analysis of the Liouville equation \eqref{LiouvillianRhopomegaDef1} for the response density written in this fashion evidences that the equations of motion of fluctuation states are 
written as a modified Sternheimer equation~\cite{mahan1980}
\be\lb{fluctuationStateEqMotion1}
\left[\omega - (\hnot-\eps_p)\right]\ket{f_p^\Phi(\omega)} = \op Q_0(\op\Phi(\omega)+\op V'[\dm'_\Phi])\ket{\psi_p} \;,
\ee
and we observe that the first-order Hamiltonian contains, apart from the perturbing field, a further term $\op V'[\dm'_\Phi]$ due to the density-dependence of $\hnot$.

\begin{figure}[t]
\includegraphics[scale=0.56]{CO_alpha+D0_vsBoxSize.eps}
\caption{\label{co_alphaStatic}(Color online) Convergence of the statical polarizability of the $CO$ with respect to the size of the computational domain.}
\end{figure}

% The interesting aspect of these equations in the perspective of the present analysis is that they allow us to present a localization argument for the fluctuation states using a procedure  
% analogous to the one depicted in section \ref{SEopenSystem}. So we split the unperturbed Hamiltonian as the sum of the kinetic $\op T$ and potential $\op V$ term and express the solution of 
% \eqref{fluctuationStateEqMotion1} in the form of a self-consistent inhomogeneous Helmholtz equation
It is easy to see that the solution of equation \eqref{fluctuationStateEqMotion1} can be casted in the form of
a set of scattering problems:
%a self-consistent inhomogeneous Helmholtz equation
\be\lb{rhoPrimeFluctuationStateDef2}
\ket{f_p^\Phi(\omega)} = \op G_H(\omega+\epsilon_p)\ket{s_p^\Phi(\omega)} \;,
\ee
with:%here in this case the source term contains further addend apart from the unperturbed potential and reads:
\be
\ket{s_p^\Phi(\omega)} =  \op V\ket{f_p^\Phi(\omega)} + \op Q_0(\op V'(\omega) + \op \Phi(\omega))\ket{\psi_p} \;.\nn
\ee
As a \emph{stringent} hypothesis, 
let us now assume that the operator $\op Q_0(\op V'(\omega) + \op \Phi(\omega))$ is, like $\op V$ such that its action
can be restricted (if suitable) on a bounded real-space domain.
%The localization character of fluctuation states can be assessed by following the same argument discussed in section \ref{SEopenSystem} for the eigenstates of the SE. In this case 
%the source term depends also from the perturbation added to the system so this analysis applies only to the class 
%of perturbing field for which $\ket{\op\Phi\psi_p}$ is localized, when projected
  %in the $\r$-representation. 
It can be seen that for each choice of $p$, the value $\omega = |\eps_p|$ represents the threshold level 
that govern the localization properties of the corresponding 
fluctuation state. Even when the above locality argument holds true, for $\omega > |\eps_p|$, 
the Helmholtz kernel behaves as a spherical wave and  gives raise to a $f_p(\omega,\r)$ which is \emph{delocalized} in the entire real-space domain.
Such delocalization is an intrinsic consequence of the value of $\omega$, and it does not depend on the particular
choice of the operators.
Only for $\omega < |\eps_p|$ the kernel contains an exponential damping factor and $f_p(\omega,\r)$ \emph{may} 
exhibit a bound-state behavior in the long range, once again provided that the above locality arguments on $\op V'$ and $\Phi$ are valid.

Such locality property of the fluctuation state can be directly confirmed for static perturbations at $\omega=0$.
Let us consider a static perturbing operator $\op\Phi_s$ that can be restricted to a finite domain. 
Any local operator, like e.g. a static electric field in the direction $\vec F$, $\Phi_s(\r) = -\vec F \cdot \r$, would satisfy such property.
We consider the perturbed hamiltonian $\hnot[\rho_{\Phi_s}] + \Phi_s$, where we indicate with $\rho_{\Phi_s}$ the GS density of the
perturbed system. We assume that the perturbation strength ($|\vec F|$ in the above exemple) is sufficiently small to be consistent with a LR 
treatment of the problem. 
In this case, the occupied eigenfunction $\ket{\psi_p^{\Phi_s}}$ of the perturbed hamiltonian 
can be extracted with traditional GS techniques, as by the hypothesis of Eq.~\eqref{SEHelmholtz1},
they would exhibit bound-state behaviour.\footnote{The fulfilling of Eq.~\eqref{SEHelmholtz1} for energies below threshold
is not \textit{per se} a guarantee of a localized $\ket{\psi}$. With such equation we demonstrate that, \emph{in the case when} a localized
solution exists, this cannot be found for energies above threshold.}

It is then easy to see that the fluctuation state at $\omega=0$ can be expressed
in terms of the corresponding bound state of the perturbed Hamiltonian, i.e.
$\ket{f_p^{\Phi_s}(0)} \simeq \op Q_0 \ket{\psi_p^{\Phi_s}}$.
For any value of $p$, the locality of the $\ket{f_p^{\Phi_s}(0)}$ therefore directly stems from the
bound-state behaviour of $\ket{\psi_p^{\Phi_s}}$. As $\omega=0$ is by hypothesis 
lower than $|\eps_p|$ this is consistent with the above observation.

@@

This fact has interesting implication in the evaluation of LR quantities.
%
% In the next section we will discuss the implication of the existence of these
% two distinct regimes on the numerical convergence of the observables computed within the LR framework. 
% Before doing this we propose an analogous analysis with the aim to understand the localization 
% properties of the excitation-based description of the LR.
% 
% The localization properties of fluctuation states allows us to assess the reliability of the linear response functional
% computed using the given basis set. 
%
Indeed, writing the response 
density in terms of fluctuation states implies that 
\be\lb{LinearResponseFunctDef2}
\braket{\delta\op O}_\Phi(\omega) = 2\mathrm{Re} \left(\sum_p \bra{\psi_p}\op O\ket{f_p^\Phi(\omega)}\right) \;.
\ee
%so the computation of the linear response functional passes through the evaluation of the scalar products indicated in \eqref{LinearResponseFunctDef2}.
Let us suppose that also the observable of interest $\op O$ exhibit the locality properties of the potentials $ \op V$, $\op \Phi$, $\op V'$.
Once again, it would be enough to consider a local operator for this condition to be met.
We observe that there exists two 
distinct regimes, on the basis of value of $\omega$. 
The first one, which we will refer to as the \emph{below threshold} regime, is realized when
$\omega$ is lower than the ionization potential $|\eps_h|$.
In this case all the fluctuation states are below their threshold level and equation \eqref{LinearResponseFunctDef2} 
is expressed in terms of genuinely localized 
quantities. The quantities $\braket{\psi_p \op O| f_p^\Phi(\omega)}$, 
being scalr products of bound-state-like objects, may thus be evaluated with computational setup that
are similar to the ones usually employed for GS calculations.
This consideration proves why, when a given computer code employs established GS numerical techniques, 
\emph{it is much easier to converge LR quantities for $\omega$ below threshold}.
The convergence in this regime is ``strong'', in the sense that the basis functions employed are able to express
wit arbitrary precision \emph{both} the states $\ket{\op O \psi_p}$ and $\ket{f_p^\Phi(\omega)}$.

@@

As an explicit example in support of these arguments we compare the absorption spectrum of a $CO$ molecule computed in 
three distinct real-space computational setups, which differs for the dimension of 
the computational domain. Results reported in figure \ref{co_spectrum} evidences that the curves are almost coincident in the below threshold regime,
whereas the effect of the computational setup are clearly visible for values of $\omega$ that exceed the threshold value.
For these ultraviolet regions, convergence might be reached for domains of much larger sizes than what presented here~\cite{baroni2008},
which are more than sufficient below threshold.

>>>>

It is easy to see from these concepts that a computational treatment that is tailored for the discretization of bound states
will not be effective for continuum states, due to the different long-range asymptotic behaviour - which in turn is a clear consequence of the 
BC of the problem. In a numerical treatment, discretizing the problem with a \emph{finite}
number of degrees of freedom, we will only have acces to \emph{pseudo-} continuum states,
whose energy values and density of states will depend on the representation of the kinetic operator
in the basis set. As a consequence, numerical eigenstates with $\epsilon > 0$ will have a value of their energy that will \emph{always}
depend on the computational setup.

A direct numerical evidence of this problem has been discussed in \cite{boffi2016}. The authors of this paper analyze the effects of a finite basis set on the 
occupied and virtual orbitals of a molecular system and establish a clear correspondence between the orbital localization character, 
and the independence of its energy with respect to the basis set. It is shown that (localized) orbitals with a bound-state behavior in the long range constitute a genuine discrete set and a finite basis is 
able to provide a reliable description of these states if the sampled domain contains their natural sizes. On the contrary, unbound (delocalized) orbitals form a continuum of states in the 
true system with open boundaries. For these class of objects the presence of a finite computational domain has a deeper impact since introduces a fictitious discretization of the energy levels. 
Furthermore, the energies of the unbound orbitals are not stable with respect to a change of the size of the computational domain and approach to zero as long as it is increased.

@@

In the contest of the LR the response density satisfies an equation of motion written in the form of a quantum Liouville operator (for example see \cite{baroni2008})
\be\lb{LiouvillianRhopomegaDef1}
\left(\omega - \Liouv\right) \dm'_\Phi(\omega) =  \commutator{\op\Phi(\omega)}{\dmnot} \;,
\ee
where $\op\Phi(\omega)$ represents the perturbing field. 
The inspection of the linear order time evolution of the eigenstates of $\hnot$ evidence that the response density acts as a transition 
operator, linking the occupied and empty subspaces of $\hnot$, and satisfies the \emph{transverse} condition  
\be\lb{RhopTransverseDef1}
\dm'_\Phi = \dm_{\perp} \doteq \dmnot\dm'_\Phi\op Q_0 + \op Q_0\dm'_\Phi\dmnot \;,
\ee
where $\op Q_0=\identity-\dmnot$ is the projector in the subspace orthogonal to the occupied states of $\hnot$. 
Due to this fact we can introduce an explicit representation of the 
response density parametrized as
\be\lb{rhoPrimeFluctuationStateDef1}
\dm'_\Phi(\omega) = \sum_p\left(\ketbra{\psi_p}{f_p^\Phi(-\omega)} + \ketbra{f_p^\Phi(\omega)}{\psi_p}\right) \;.
\ee
Here we have introduced the set of $\omega$-dependent \emph{fluctuation states} $\{\ket{f_p^\Phi(\omega)}\}$, 
that have values in the \emph{unoccupied} subspace of $\hnot$. 
An analysis of the Liouville equation \eqref{LiouvillianRhopomegaDef1} for the response density written in this fashion evidences that the equations of motion of fluctuation states are 
written as a modified Sternheimer equation~\cite{mahan1980}
\be\lb{fluctuationStateEqMotion1}
\left[\omega - (\hnot-\eps_p)\right]\ket{f_p^\Phi(\omega)} = \op Q_0(\op\Phi(\omega)+\op V'[\dm'_\Phi](\omega))\ket{\psi_p} \;,
\ee
and we observe that the first-order Hamiltonian contains, 
apart from the perturbing field, a further term $\op V'[\dm'_\Phi]$ due to the density-dependence of $\hnot$.
% The interesting aspect of these equations in the perspective of the present analysis is that they allow us to present a localization argument for the fluctuation states using a procedure  
% analogous to the one depicted in section \ref{SEopenSystem}. So we split the unperturbed Hamiltonian as the sum of the kinetic $\op T$ and potential $\op V$ term and express the solution of 
% \eqref{fluctuationStateEqMotion1} in the form of a self-consistent inhomogeneous Helmholtz equation
It is easy to see that the solution of equation \eqref{fluctuationStateEqMotion1} can be casted in the form of
a set of scattering problems:
%a self-consistent inhomogeneous Helmholtz equation
\be\lb{rhoPrimeFluctuationStateDef2}
\ket{f_p^\Phi(\omega)} = \op G_H(\omega+\epsilon_p)\left(\op V\ket{f_p^\Phi(\omega)} + \ket{s_p^\Phi(\omega)} \right)\;,
\ee
with%:here in this case the source term contains further addend apart from the unperturbed potential and reads:
$%\be
\ket{s_p^\Phi(\omega)} = \op Q_0(\op V'[\dm'_\Phi](\omega) + \op \Phi(\omega))\ket{\psi_p}
$. %\ee

It can be seen that for each choice of $p$, the value $\omega = |\eps_p|$ represents the threshold level 
that govern the localization properties of the corresponding 
fluctuation state.
As a \emph{stringent} hypothesis, 
let us now assume that the ket $\ket{s_p^\Phi(\omega)}$ is localized, namely its evaluation can be
restricted (if suitable) on a bounded real-space domain.
%The localization character of fluctuation states can be assessed by following the same argument discussed in section \ref{SEopenSystem} for the eigenstates of the SE. In this case 
%the source term depends also from the perturbation added to the system so this analysis applies only to the class 
%of perturbing field for which $\ket{\op\Phi\psi_p}$ is localized, when projected
  %in the $\r$-representation. 
Even when such locality argument holds true, for $\omega > |\eps_p|$, 
the Helmholtz kernel behaves as a spherical wave and  gives raise to a $f_p(\omega,\r)$ which is \emph{delocalized} in the entire real-space domain.
Such delocalization is an intrinsic consequence of the value of $\omega$, and it does not depend on the particular
choice of the operators.
Only for $\omega < |\eps_p|$ the kernel contains an exponential damping factor and $f_p(\omega,\r)$ \emph{may} 
exhibit a bound-state behavior in the long range, once again provided that the above locality arguments on $\ket{s_p^\Phi(\omega)}$ 
-- hence on $\op V'$ and $\Phi$ -- are valid.

Such locality property of the fluctuation state can be directly confirmed for static perturbations at $\omega=0$.
Let us consider a static perturbing operator $\op\Phi_s$ that can be restricted to a finite domain. 
Any local operator, like e.g. a static electric field in the direction $\mathbf F$, $\Phi_s(\r) = -\mathbf F \cdot \r$, would satisfy such property.
We consider the perturbed hamiltonian $\hnot[\rho_{\Phi_s}] + \Phi_s$, where we indicate with $\rho_{\Phi_s}$ the GS density of the
perturbed system. We assume that the perturbation strength ($F$ in the above exemple) is sufficiently small to be consistent with a LR 
treatment of the problem. 
In this case, the occupied eigenfunction $\ket{\psi_p^{\Phi_s}}$ of the perturbed hamiltonian 
can be extracted with traditional GS techniques, as by the hypothesis of Eq.~\eqref{SEHelmholtz1},
they would exhibit bound-state behaviour.\footnote{The fulfilling of Eq.~\eqref{SEHelmholtz1} for energies below threshold
is not \textit{per se} a guarantee of a localized $\ket{\psi}$. With such equation we demonstrate that, \emph{in the case when} a localized
solution exists, this cannot be found for energies above threshold.}
\begin{figure}[t]
\includegraphics[scale=0.56]{CO_alpha+D0_vsBoxSize.eps}
\caption{\label{co_alphaStatic}(Color online) Convergence of the statical polarizability of the $CO$ with respect to the size of the computational domain.}
\end{figure}

From the linearized Schroedinger Equation of the perturbed problem, it is easy to see that the fluctuation state at $\omega=0$ can be expressed
in terms of the corresponding bound state of the perturbed Hamiltonian, i.e.
$\ket{f_p^{\Phi_s}(0)} \simeq \op Q_0 \ket{\psi_p^{\Phi_s}}$.
For any value of $p$, the locality of the $\ket{f_p^{\Phi_s}(0)}$ therefore directly stems from the
bound-state behaviour of $\ket{\psi_p^{\Phi_s}}$. As $\omega=0$ is by hypothesis 
lower than $|\eps_p|$, this is consistent with the above considerations.

@@

This fact has interesting implication in the evaluation of LR quantities.
%
% In the next section we will discuss the implication of the existence of these
% two distinct regimes on the numerical convergence of the observables computed within the LR framework. 
% Before doing this we propose an analogous analysis with the aim to understand the localization 
% properties of the excitation-based description of the LR.
% 
% The localization properties of fluctuation states allows us to assess the reliability of the linear response functional
% computed using the given basis set. 
%
Indeed, writing the response 
density in terms of fluctuation states implies that 
\be\lb{LinearResponseFunctDef2}
\braket{\delta\op O}_\Phi(\omega) = 2\mathrm{Re} \left(\sum_p \bra{\psi_p}\op O\ket{f_p^\Phi(\omega)}\right) \;.
\ee
%so the computation of the linear response functional passes through the evaluation of the scalar products indicated in \eqref{LinearResponseFunctDef2}.
Let us suppose that also the observable of interest $\op O$ exhibit the locality properties of the potentials $ \op V$, $\op \Phi$, $\op V'$.
Once again, it would be enough to consider a local operator for this condition to be met.
We observe that there exists two 
distinct regimes, on the basis of value of $\omega$. 
The first one, which we will refer to as the \emph{below threshold} regime, is realized when
$\omega$ is lower than the ionization potential $|\eps_h|$.
In this case all the fluctuation states are below their threshold level and equation \eqref{LinearResponseFunctDef2} 
is expressed in terms of genuinely localized 
quantities. The quantities $\braket{\psi_p \op O| f_p^\Phi(\omega)}$, 
being scalr products of bound-state-like objects, may thus be evaluated with computational setup that
are similar to the ones usually employed for GS calculations.
This consideration proves why, when a given computer code employs established GS numerical techniques, 
\emph{it is much easier to converge LR quantities for $\omega$ below threshold}.
The convergence in this regime is ``strong'', in the sense that the basis functions employed are able to express 
with arbitrary precision \emph{both} the states $\ket{\op O \psi_p}$ and $\ket{f_p^\Phi(\omega)}$.
This is of course true even when the above states are only implicitly expressed by the employed LR treatment,
 as e.g. in the case of Krylov spaces generated from the states $\ket{\op\Phi(\omega) \psi_p}$ (see.e.g. \cite{baroni2008,linlinKPM}).

@@

As an explicit example in support of these arguments we compare the absorption spectrum of a $CO$ molecule computed in 
three distinct real-space computational setups, which differs for the dimension of 
the computational domain. Results reported in figure \ref{co_spectrum} evidences that the curves are almost coincident in the below threshold regime,
whereas the effect of the computational setup are clearly visible for values of $\omega$ that exceed the threshold value.
For these ultraviolet regions, convergence might be reached for domains of much larger sizes than what presented here~\cite{baroni2008},
which in turn reveal more than sufficient below threshold.
